### 2013-09-22 latex/paper-PLOS_ONE-revised.tex
First draft with all reviewer comments addressed.

\usepackage{mathptmx} 
\usepackage{graphicx}
\usepackage{lscape}
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage{cite}
\usepackage{float}
\renewcommand{\bottomfraction}{1.}
\renewcommand{\topfraction}{1.}
\renewcommand{\textfraction}{0.}
\renewcommand{\floatpagefraction}{1.}

@@

@@

As RSA calculations are based on the Gly-X-Gly framework, we excluded all chain terminating residues from both the empirical and the theoretical analysis. Even with our improved ASA normalization values, then, chain-terminating residues may still display $\text{RSA}>1$. We therefore recommend that future analyses making use of RSA similarly exclude any chain-terminating residues, as their RSA estimates will not be precise. Suitable normalization values for chain-terminating residues are not available at present.

@@

In many applications, specifically in the context of sequence evolution, RSA is treated as a site-specific property that is invariant under mutation. While RSA values of homologous structures tend to be strongly correlated \cite{RostSander1994,Zhouetal2009}, individual sites, in particular exposed ones, can show substantial RSA variability \cite{RostSander1994}. In this context, it is important to note that one potential source of RSA variability in previous studies was RSA normalization. For example, Ref.~\cite{RostSander1994} used the Rose scale, which differs quite substantially from the scale we propose here. In particular, the corrections we propose to the Rose scale range from 4\% (for Leu) to 18\% (for Asp), and are approximately uniformly distributed in that range over the 20 amino acids. Thus, many scenarios are possible under which a substitution that might not change RSA under our scale might change it by over 10\% under the Rose scale. A detailed analysis of RSA variability among homologous structures is beyond the scope of this work, but should be undertaken in future work.

@@

\subsection*{Data and code deposition}
All results and all computer code used to generate these results have been deposited to GitHub.com (https://github.com/mtien/RSA-normalization-values). This inlcudes maximum observed ASA values (both empirical and theoretical) as a function of backbone dihedral angles.

@@

\bibliography{bibliography}

\newpage

>>>>

\usepackage{mathptmx} 
\usepackage{graphicx}
\usepackage{lscape}
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage{cite}
\usepackage{float}
\usepackage{color}
\renewcommand{\bottomfraction}{1.}
\renewcommand{\topfraction}{1.}
\renewcommand{\textfraction}{0.}
\renewcommand{\floatpagefraction}{1.}

@@

Our results are broadly consistent with a recent paper by Singh and Ahmad~\cite{SinghAhmad2009}. These authors did an extensive empirical survey of ASA values in tripeptides from PDB structures. They found that the highest observed ASA values were found in loops and turns, not in the extended conformation used by Miller \emph{et al.}. Their highest ASA values are generally consistent with ours. Further, Singh and Ahmad found that the highest observed ASA values were dependent on the neighboring residues around the focal residue. Finally, Singh and Ahmad showed that for RSA prediction from primary sequence, prediction accuracy could be improved by approximately 10\% if ASA values were normalized by (neighbor-dependent) highest observed ASA values rather than by ASA values observed in the extended conformation~\cite{SinghAhmad2009}. Our work serves as a useful complement to their work, by (i) providing, through molecular modeling, highest \emph{possible} ASA values rather than just highest \emph{observed} ASA values, by (ii) providing highest observed and highest possible ASA values as a function of backbone dihedral angle, and by (iii) demonstrating that improved RSA normalization yields empirical hydrophobicity scales that are more similar to experimentally measured ones.

In our modeling approach, we calculated ASA values for Gly-X-Gly tripeptides. Other authors have considered normalizations based on Ala-X-Ala tripeptides \cite{Ahmadetal2003,NguyenRajapakse2005} or even neighbor-specific normalizations (i.e., a different normalization for each specific tripeptide \cite{SinghAhmad2009}). We chose Gly-X-Gly tripeptides because we wanted to calculate the highest possible ASA values of tripeptides, and glycines will generally occlude less solvent than alanines. From a practical perspective, we prefer a simple normalization scheme, and hence highest possible ASA values are attractive to us. However, for certain applications, it may be the case that neighbor-specific or backbone-specific normalizations are preferable. Singh and Ahmad \cite{SinghAhmad2009} provided neighbor-specific normalization values, but didn't control for backbone angles. We have shown here that maximum ASA values depend substantially on backbone angles (e.g.\ Fig.~\ref{fig:heatrama}), and we provide both highest observed and highest possible ASA values as a function of backbone angles (see ``Data and code availability'' in Methods). It is not known at this time whether neighbor-dependent or backbone-dependent normalization is preferable, and the answer may depend on the specific application. In principle, one could also normalize by both neighboring amino acids and backbone dihedral angles. A modeling approach such as ours could be employed to calculate the highest possible ASA values for any tripeptide in any conformation. The computational resources required would be substantial, however, since we would have to model 400 times more tripeptides than we did for the present work.

@@

As our RSA calculations are based on ASAs of tripeptides, we excluded all chain terminating residues from both the empirical and the theoretical analysis. Even with our improved ASA normalization values, then, chain-terminating residues may still display $\text{RSA}>1$. We therefore recommend that future analyses making use of RSA similarly exclude any chain-terminating residues, as their RSA estimates will not be precise. Suitable normalization values for chain-terminating residues are not available at present.

@@

In many applications, specifically in the context of sequence evolution, RSA is treated as a site-specific property that is invariant under mutation. While RSA values of homologous structures tend to be strongly correlated \cite{RostSander1994,Zhouetal2009}, individual sites, in particular exposed ones, can show substantial RSA variability \cite{RostSander1994}. In this context, we would like to emphasize that one potential source of RSA variability in previous studies was RSA normalization. For example, Ref.~\cite{RostSander1994} used the Rose scale, which differs quite substantially from the scale we propose here. In particular, the corrections we propose to the Rose scale range from 4\% (for Leu) to 18\% (for Asp), and are approximately uniformly distributed in that range over the 20 amino acids. Thus, one can envision scenarios under which a substitution that might not change RSA under our scale might change it by over 10\% under the Rose scale.
At the same time, we have to realize that RSA can show variability even in the absence of mutation, in particular for exposed residues. A residue in a surface loop will undergo thermodynamic fluctuations, and its solvent exposure state will vary over time as neighboring residues move closer in or further out. By contrast, a residue in the core will likely remain solvent-occluded at all times. To obtain a reliable RSA value for a surface residue, one would thus ideally calculate an average over a thermodynamic ensemble of structures. A detailed analysis of RSA variability under thermodynamic fluctuations and among homologous structures is beyond the scope of this work but should be undertaken in the future.

@@

\subsection*{Data and code availability}
All results and all computer code used to generate these results have been deposited to GitHub.com (https://github.com/mtien/RSA-normalization-values). This inlcudes maximum observed ASA values (both empirical and theoretical) as a function of backbone dihedral angles.

@@

%\bibliography{bibliography}

\begin{thebibliography}{10}
\providecommand{\url}[1]{\texttt{#1}}
\providecommand{\urlprefix}{URL }
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi
\providecommand{\bibAnnoteFile}[1]{%
  \IfFileExists{#1}{\begin{quotation}\noindent\textsc{Key:} #1\\
  \textsc{Annotation:}\ \input{#1}\end{quotation}}{}}
\providecommand{\bibAnnote}[2]{%
  \begin{quotation}\noindent\textsc{Key:} #1\\
  \textsc{Annotation:}\ #2\end{quotation}}
\providecommand{\eprint}[2][]{\url{#2}}

\bibitem{Chothia1976}
Chothia C (1976) The nature of the accessible and buried surfaces in proteins.
\newblock J Mol Biol : 1--14.
\bibAnnoteFile{Chothia1976}

\bibitem{Rose1985}
Rose GD, Geselowitz AR, Lesser GJ, Lee RH, Zehfus MH (1985) Hydrophobicity of
  amino acid residues in globular proteins.
\newblock Science 229: 834--838.
\bibAnnoteFile{Rose1985}

\bibitem{Miller1987}
Miller S, Janin J, Lesk AM, Chothia C (1987) Interior and surface of monomeric
  proteins.
\newblock J Mol Biol 196: 641--656.
\bibAnnoteFile{Miller1987}

\bibitem{Moelbert2004}
Moelbert S, Emberly E, Tang C (2004) Correlation between sequence
  hydrophobicity and surface-exposure pattern of database proteins.
\newblock Prot Sci : 752--762.
\bibAnnoteFile{Moelbert2004}

\bibitem{Shaytan2009}
Shaytan AK, Shaitan KV, Khokhlov AR (2009) Solvent accessible surface area of
  amino acid residues in globular proteins: Correlations of apparent transfer
  free engergies with experimental hydrophobicity scales.
\newblock Biomacromolecules 10: 1224--1237.
\bibAnnoteFile{Shaytan2009}

\bibitem{Goldmanetal1998}
Goldman N, Thorne JL, Jones DT (1998) Assessing the impact of secondary
  structure and solvent accessibility on protein evolution.
\newblock Genetics 149: 445--458.
\bibAnnoteFile{Goldmanetal1998}

\bibitem{Bloometal2006}
Bloom JD, Drummond DA, Arnold FH, Wilke CO (2006) Structural determinants of
  the rate of protein evolution in yeast.
\newblock Mol Biol Evol 23: 1751--1761.
\bibAnnoteFile{Bloometal2006}

\bibitem{FranzosaXia2009}
Franzosa EA, Xia Y (2009) Structural determinants of protein evolution are
  context-sensitive at the residue level.
\newblock Mol Biol Evol 26: 2387--2395.
\bibAnnoteFile{FranzosaXia2009}

\bibitem{Zhouetal2009}
Zhou T, Weems M, Wilke CO (2009) Translationally optimal codons associate with
  structurally sensitive sites in proteins.
\newblock Mol Biol Evol 26: 1571--1580.
\bibAnnoteFile{Zhouetal2009}

\bibitem{FranzosaXia2012}
Franzosa EA, Xia Y (2012) Independent effects of protein core size and
  expression on structure-evolution relationships at the residue level.
\newblock PLoS One 7: e46602.
\bibAnnoteFile{FranzosaXia2012}

\bibitem{Scherrer2012}
Scherrer MP, Meyer AG, Wilke CO (2012) Modeling coding-sequence evolution
  within the context of residue solvent accessibility.
\newblock BMC Evol Biol 12: 179.
\bibAnnoteFile{Scherrer2012}

\bibitem{MeyerWilke2012}
Meyer AG, Wilke CO (2012) Integrating sequence variation and protein structure
  to identify sites under selection.
\newblock Mol Biol Evol .
\bibAnnoteFile{MeyerWilke2012}

\bibitem{ConantStadler2009}
Contant GC, Stadler PF (2009) Solvent exposure imparts similar selective
  pressures across a range of yeast proteins.
\newblock Mol Biol Evol 26: 1155--1161.
\bibAnnoteFile{ConantStadler2009}

\bibitem{RostSander1994}
Rost B, Sander C (1994) Conservation and prediction of solvent accessibility in
  protein families.
\newblock Proteins 20: 216--226.
\bibAnnoteFile{RostSander1994}

\bibitem{Pollastrietal2002}
Pollastri G, Baldi P, Fariselli P, Casadio R (2002) Prediction of coordination
  number and relative solvent accessibility in proteins.
\newblock Proteins 47: 142--153.
\bibAnnoteFile{Pollastrietal2002}

\bibitem{KimPark2004}
Kim H, Park H (2004) Prediction of protein relative solvent accessibility with
  support vector machines and long-range interaction {3D} local descriptor.
\newblock Proteins 54: 557--562.
\bibAnnoteFile{KimPark2004}

\bibitem{Adamczaketal2004}
Adamczak R, Porollo A, Meller J (2004) Accurate prediction of solvent
  accessibility using neural networks-based regression.
\newblock Proteins: Structure, Function, and Bioinformatics 56: 753--767.
\bibAnnoteFile{Adamczaketal2004}

\bibitem{NguyenRajapakse2005}
Nguyen MN, Rajapakse JC (2005) Prediction of protein relative solvent
  accessibility with a two-stage {SVM} approach.
\newblock Proteins 59: 30--37.
\bibAnnoteFile{NguyenRajapakse2005}

\bibitem{Petersenetal2009}
Petersen B, {Nordahl Petersen} T, Andersen P, Nielsen M, Lundegaard C (2009) A
  generic method for assignment of reliability scores applied to solvent
  accessibility predictions.
\newblock BMC Struct Biol 9: 51.
\bibAnnoteFile{Petersenetal2009}

\bibitem{SinghAhmad2009}
Singh H, Ahmad S (2009) Context dependent reference states of solvent
  accessibility derived from native protein sturctures and assessed by
  predictability analysis.
\newblock BMC Struct Biol 9: 25.
\bibAnnoteFile{SinghAhmad2009}

\bibitem{Levy2010}
Levy ED (2010) A simple definition of structural regions in proteins and its
  use in analyzing interface evolution.
\newblock J Mol Biol 403: 660--670.
\bibAnnoteFile{Levy2010}

\bibitem{Chengetal2006}
Cheng J, Sweredoski MJ, Baldi P (2006) {DOMpro:} protein domain prediction
  using profiles, secondary structure, relative solvent accessibility, and
  recursive neural networks.
\newblock Data Mining and Knowledge Discovery 13: 1--10.
\bibAnnoteFile{Chengetal2006}

\bibitem{ChenZhou2005}
Chen H, Zhou HX (2005) Prediction of solvent accessibility and sites of
  deleterious mutations from protein sequence.
\newblock Nucl Acids Res 33: 3193--3199.
\bibAnnoteFile{ChenZhou2005}

\bibitem{WangDunbrack2003}
Wang G, Dunbrack RL (2003) {PISCES:} a protein sequence culling server.
\newblock Bioinformatics 19: 1589--1591.
\bibAnnoteFile{WangDunbrack2003}

\bibitem{Morrisetal1992}
Morris AL, MacArthur MW, Hutchinson EG, Thornton JM (1992) Stereochemical
  quality of protein structure coordinates.
\newblock Proteins 12: 345--364.
\bibAnnoteFile{Morrisetal1992}

\bibitem{Wolfenden1981}
Wolfenden R, Anderson L, Cullis PM, Southgate CCB (1981) Affinities of amino
  acid side chains for solvent water.
\newblock Biochemistry : 849--855.
\bibAnnoteFile{Wolfenden1981}

\bibitem{Kyte1981}
Kyte J, Doolittle RF (1982) A simple method for displaying the hydropathic
  character of a protein.
\newblock J Mol Biol 157: 105--132.
\bibAnnoteFile{Kyte1981}

\bibitem{Fauchere1983}
Fauchere JL, Pliska VE (1983) Hydrophobic parameters of amino-acid side chains
  from the partitioning of {N-acetyl-amino-acid} amides.
\newblock Eur J Med Chem 18: 369-375.
\bibAnnoteFile{Fauchere1983}

\bibitem{Radzicka1988}
Radzicka A, Wolfenden R (1988) Comparing the polarities of the amino acids:
  side-chain distribution coefficients between the vapor phase, cyclohexane,
  1-octanol, and neutral aqueous solution.
\newblock Biochem 27: 1664--1670.
\bibAnnoteFile{Radzicka1988}

\bibitem{MacCallum2007}
MacCallum JL, Bennett WFD, Tieleman DP (2007) Partitioning of amino acid side
  chains into lipid bilayers: results from computer simulations and comparison
  to experiment.
\newblock J Gen Physiol 129: 371--377.
\bibAnnoteFile{MacCallum2007}

\bibitem{Moon2011}
Moon CP, Fleming KG (2011) Side chain hydrophobicity scales derived from
  transmembrane protein folding into lipid bilayers.
\newblock Proc Natl Acad Sci USA 108: 10174--10177.
\bibAnnoteFile{Moon2011}

\bibitem{Wimley1996}
Wimley WC, Creamer TP, White SH (1996) Solvation energies of amino acid side
  chains and backbone in a family of host-guest pentapeptides.
\newblock Biochem 35: 5109--5124.
\bibAnnoteFile{Wimley1996}

\bibitem{Ahmadetal2003}
Ahmad S, Gromiha MM, Sarai A (2003) Real-value prediction of solvent
  accessibility from amino acid sequence.
\newblock Proteins 50: 629--635.
\bibAnnoteFile{Ahmadetal2003}

\bibitem{Kabsch1983}
Kabsch W, Sander C (1983) Dictionary of protein secondary structure: pattern
  recognition of hydrogen-bonded and geometrical features.
\newblock Biopolymers 22: 2577--2637.
\bibAnnoteFile{Kabsch1983}

\bibitem{LeeRichards1971}
Lee B, Richards FM (1971) The interpretation of protein structures: estimation
  of static accessibility.
\newblock J Mol Biol 55: 379--400.
\bibAnnoteFile{LeeRichards1971}

\bibitem{KawashimaKanehisa2000}
Kawashima S, Kanehisa M (2000) {AAindex:} amino acid index database.
\newblock Nucleic Acids Res 28: 374.
\bibAnnoteFile{KawashimaKanehisa2000}

\bibitem{TomiiKanehisa1996}
Tomii K, Kanehisa M (1996) Analysis of amino acid indices and mutation matrices
  for sequence comparison and structure prediction of proteins.
\newblock Protein Eng 9: 27--36.
\bibAnnoteFile{TomiiKanehisa1996}

\bibitem{Kawashimaetal2008}
Kawashima S, Pokarowski P, Pokarowska M, Kolinski A, Katayama T, et~al. (2008)
  {AAindex:} amino acid index database, progress report 2008.
\newblock Nucleic Acids Res 36: D202--D205.
\bibAnnoteFile{Kawashimaetal2008}

\bibitem{Laskowskietal1993}
Laskowski RA, MacArthur MW, Moss DS, Thornton JM (1993) {PROCHECK} -- a program
  to check the stereochemical quality of protein structures.
\newblock J App Cryst 26: 283--291.
\bibAnnoteFile{Laskowskietal1993}

\bibitem{Tienetal2013}
Tien MZ, Sydykova DK, Meyer AG, Wilke CO (2013) {PeptideBuilder:} a simple
  {Python} library to generate model peptides.
\newblock PeerJ 1: e80.
\bibAnnoteFile{Tienetal2013}

\end{thebibliography}
\newpage
